Illustrating imputation of missing data using the Titanic dataset

### 2024-05-11 text/pending/survival-lasso.Rmd
First full draft of survival-lasso

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(glmnet)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "whas500-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

Here is a brief description of the whas100 dataset, taken from the [data dictionary][sim3] on my github site.

@@

```{r}
vnames <- names(dd$vars)
f2 <- "whas500.dat"
w0 <- read_table(
  file=paste0(f0, f2), 
  col_names=c(vnames),
  col_types="nnnnnnnnnnnnnnncccnnnn")
head(w0)
```

@@

```{r}
dd$vars[["id"]]
summary(w0$id)

dd$vars[["age"]]
summary(w0$age)

dd$vars[["gender"]]
count(w0, gender)

dd$vars[["sysbp"]]
summary(w0$sysbp)

dd$vars[["diasbp"]]
summary(w0$diasbp)

dd$vars[["bmi"]]
summary(w0$bmi)

dd$vars[["cvd"]]
count(w0, cvd)

dd$vars[["afb"]]
count(w0, afb)

dd$vars[["sho"]]
count(w0, sho)

dd$vars[["chf"]]
count(w0, chf)

dd$vars[["av3"]]
count(w0, av3)

dd$vars[["miord"]]
count(w0, miord)

dd$vars[["mitype"]]
count(w0, mitype)

dd$vars[["year"]]
count(w0, year)

dd$vars[["los"]]
summary(w0$los)

dd$vars[["dstat"]]
count(w0, dstat)

dd$vars[["lenfol"]]
summary(w0$lenfol)

dd$vars[["fstat"]]
count(w0, fstat)
```

@@

You can find the lasso regression model in the glmnet package. One important difference with glmnet is that it does not use a formula to define your model. Instead you provide matrices for your dependent variable and your independent variables.

>>>>

I've followed the [glmnet vignette for Cox regression][tay1] closely. You might find that explanation to be clearer than what I present below.

[tay1]: https://glmnet.stanford.edu/articles/Coxnet.html

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(glmnet)
library(glue)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
g0 <- "https://raw.githubusercontent.com/pmean/"
r0 <- "datasets/master/"
s1 <- "yaml-functions.R"
f1 <- "whas500.yaml"

source(paste0(g0, r0, s1))
dd <- read_yaml(paste0(g0, r0, f1))
vnames <- extract_names(dd)
vlabels <- extract_labels(dd)
vscales <- extract_scales(dd)
```

Here is a brief description of the whas500 dataset, taken from the [data dictionary][sim3] on my github site.

@@

```{r}
vnames <- names(dd$vars)
f2 <- "whas500.dat"
w0 <- read_table(
  file=paste0(g0, r0, f2), 
  col_names=c(vnames),
  col_types="nnnnnnnnnnnnnnncccnnnn")
head(w0)
```

@@

```{r}
categorical <- 
  (vscales=="binary") |
  (vscales=="nominal") |
  (vscales=="ordinal")
for (i_vars in which(categorical)) {
  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
  w0 %>%
    count(.data[[vnames[i_vars]]]) %>%
    data.frame %>%
    print
  cat("\n")
}
```

```{r}
continuous <- 
  (vscales=="interval") |
  (vscales=="ratio")
for (i_vars in which(continuous)) {
  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
  print(summary(w0[vnames[i_vars]]))
  cat("\n")
}
```

@@

You can find the lasso regression model in the glmnet package. One important difference with glmnet is that it does not use a formula to define your model. Instead you provide matrices for your dependent variable and your independent variables. In a Cox regression model, the dependent matrix has a column for time and a second column with an indicator variable (1 means the event occurred and 0 means censoring).

```{r}
ymat <- cbind(w0$lenfol, w0$fstat)
dimnames(ymat)[2] <- list(c("time", "status"))
xmat <- as.matrix(w0[ , 8:14])
dimnames(xmat)[2] <- list(vnames[8:14])
lasso_fit <- glmnet(xmat, ymat, family="cox")
plot(lasso_fit, label=TRUE)
plot(lasso_fit, xvar="norm", label=TRUE)
plot(lasso_fit, xvar="lambda", label=TRUE)
plot(lasso_fit, xvar="dev", label=TRUE)
```

```{r}
cv_fit <- cv.glmnet(xmat, ymat, family = "cox", type.measure="mae")
plot(cv_fit)
plot(lasso_fit, xvar="lambda")
abline(v=log(cv_fit$lambda.1se))
cv_fit$lambda.1se
coef(lasso_fit, cv_fit$lambda.1se)
```

```{r}
cv_fit$lambda.min
plot(lasso_fit, xvar="lambda")
abline(v=log(cv_fit$lambda.min))
coef(lasso_fit, cv_fit$lambda.min)
```

### 2024-05-11 text/pending/titanic-imputation.Rmd
First full draft of survival-lasso

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(mice)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "titanic-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

The Titanic was a large cruise ship, the biggest of its kind in 1912. It was thought to be unsinkable, but when it set sail from England to American in its maiden voyage, it struck an iceberg and sank, killing many of the passengers and crew. You can get fairly good data on the characteristics of passengers who died and compare them to those that survived. The data indicate a strong effect due to age and gender, representing a philosophy of "women and children first" that held during the boarding of life boats. Let's look at the effect of age on survival using a logistic regression model.

@@

```{r}
f2 <- "titanic_v00.txt"
ti0 <- 
  data.frame(
    read_tsv(
      file=paste0(f0, f2),
      col_types="ccncn"))
names(ti0) <- tolower(names(ti0))
head(ti0)
```

@@

```{r}
summary(ti0$age)
count(ti0, pclass)
count(ti0, sex)
count(ti0, survived)
```

Notice the large number of missing values for age. The first three passegers with missing ages are

```{r}
missing_rows <- 13:15
ti0 %>% slice(missing_rows)
```

@@

```{r}
ti1 <- select(ti0, -name)
ti_mice <- mice(ti1)
```

@@

```{r}
ti_with <- with(
  ti_mice, 
  glm(
    survived~pclass+rcs(age)+sex, 
    family=binomial)) 
```

@@

```{r}
ti_pool <- pool(ti_with)
summary(ti_pool)
log_or <- summary(ti_pool)[8, 2]
```

@@

```{r}
se <- summary(ti_pool)[8, 3]
or <- exp(log_or)
lo <- exp(log_or-1.96*se)
hi <- exp(log_or+1.96*se)
paste0(round(or, 3), " (", round(lo, 3), ",", round(hi, 3), ")")
```

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(glue)
library(mice)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
g0 <- "https://raw.githubusercontent.com/pmean/"
r0 <- "datasets/master/"
f1 <- "titanic.yaml"
s1 <- "yaml-functions.R"
source(paste0(g0, r0, s1))
dd <- read_yaml(paste0(g0, r0, f1))
vnames <- extract_names(dd)
vlabels <- extract_labels(dd)
```

@@

```{r, echo=FALSE}
f2 <- "titanic_v00.txt"
raw_data <- 
  data.frame(
    read_tsv(
      file=paste0(g0, r0, f2),
      col_types="ccncn"))
# Re-arranging columns just for convenient display
o <-  c(1, 3, 2, 4, 5)
ti0 <- select(raw_data, o)
vnames <- vnames[o]
head(ti0)
```

@@

```{r, echo=FALSE}
cat(vnames[1], ": ", vlabels[1], "\n")
summary(ti0$Age)
cat(vnames[2], ": ", vlabels[2], "\n")
count(ti0, PClass)
cat(vnames[3], "\n")
count(ti0, Sex)
cat(vnames[4], "\n")
cat(as.yaml(dd$vars$Survived$values), "\n")
count(ti0, Survived)
```

I have hidden the R code up to this point, as it is mundane and not of great interest. I will show the R code and output for the rest of the analysis.

Notice the large number of missing values for age. The first three passengers with missing ages are

```{r}
missing_rows <- which(is.na(ti0$Age))[1:3]
ti0 %>% slice(missing_rows)
```

@@

```{r}
ti1 <- select(ti0, -Name)
ti1$i_female <- as.numeric(ti1$Sex=="female")
ti1 <- select(ti1, -Sex)
ti_mice <- mice(ti1)
```

@@

```{r}
ti_with <- with(
  ti_mice, 
  glm(
    Survived~PClass+rcs(Age, 3)+i_female, 
    family=binomial)) 
```

@@

```{r}
ti_pool <- pool(ti_with)
ti_summary <- summary(ti_pool)
ti_summary
```

@@

```{r}
female <- ti_summary$term=="i_female"
log_or <- ti_summary[female, "estimate"]
se <- ti_summary[female, "std.error"]
or <- round(exp(log_or), 1)
lo <- round(exp(log_or-1.96*se), 1)
hi <- round(exp(log_or+1.96*se), 1)
glue("Odds ratio for females is {or}, 95% CI ({lo},{hi})")
```
